Effect of quality-adjusted production function inputs

### 2015-07-08 Articles/Article_EcolEcon_2015/EcolEcon_2015.Rnw
Rearranging text into paper order.

# Tell whether to re-run the analysis to create the models.
# This file will use previously-saved models to generate
# the paper if set to FALSE, greatly reducing time.
# If something changes in the data or in the code that
# performs the analysis, set 
# rerunModels = TRUE
# once to re-generate the information needed to 
# create the paper.
# ********************************************************
rerunModels=TRUE
# ********************************************************
opts_chunk$set(
  # ********************************************************
  eval=TRUE,
  # ********************************************************
  # tikz allows LaTeX code in graphical output. 
  # E.g., "$y$" for a variable name in the legend for a graph.  
  # dev='tikz',
  tidy=FALSE,
  comment=NA,
  # Tells whether to cache output from chunks are saved. Cacheing saves time. 
  # However, references to LaTeX tables DO NOT WORK if cacheing is turned on.
  cache=FALSE,
  warning=FALSE,  #Tells whether to show warnings in the output.
  # Turns off messages for all chunks. 
  # Set TRUE on an individual chunk to see it.
  message=FALSE, 
  # Tells whether to echo code for all chunks. 
  # Set TRUE on an individual chunk to see its code.
  echo=FALSE,
  # Sets all table captions to the top.
  xtable.caption.placement="top"
)

@@

%------------------------------
\subsubsection{Exponential Production Function} 
\label{sec:exp}
%------------------------------

We define an exponential-only reference model for economic growth
in Equation~\ref{eq:exponential-only-model}.
%
\begin{equation} \label{eq:exponential-only-model}
  y = \theta A \; ; \; A \equiv \mathrm{e}^{\lambda(t-t_0)} \; .
\end{equation}
%
The reference model is a degenerate case of the Cobb-Douglas production function
wherein all factor shares are zero~($\alpha_1 = \alpha_2 = \alpha_3 = 0$)
and the constant returns to scale constraint is not respected.

We expect that the reference model will have a larger fitted Solow residual term than
the Cobb-Douglas and CES models, because no factors of production are
included in the reference model to drive growth.
Indeed, in the reference model, 
all growth is attributed to the Solow residual.

@@

  models <- list()
  for(country in countries){
    for (flavor in flavors){
      fnoE <- keep_summands(formulas[[country]][[flavor]], -3) # Drop the energy term (3rd term on RHS)
      fwithE <- formulas[[country]][[flavor]]
      # CD models
      models[[country]][[flavor]][["noE"]][["CD"]]   <- cdModel(fnoE, data = historicalData[[country]])
      models[[country]][[flavor]][["withE"]][["CD"]] <- cdModel(fwithE, data = historicalData[[country]])
      # CES models
      models[[country]][[flavor]][["noE"]][["CES"]][["kl"]] <- cesModel(fnoE, 
                                                                        data = historicalData[[country]], 
                                                                        nest=c(1,2))
      nestsWithE <- list(kle=c(1,2,3), lek=c(2,3,1), ekl=c(3,1,2))
      models[[country]][[flavor]][["withE"]][["CES"]] <- lapply(nestsWithE, function(nest){
          cesModel(formula=fwithE, data=historicalData[[country]], nest=nest)
        })
    }      
  }
  # Save models to disk.
  saveRDS(models, file=fileName)
}
# Put all exponential model coefficients into a data frame
AllCoeffsExp <- leaf_apply(modelsExp, class = c("lm"), f = function(modExp, id){
  # Parse the id to find details of this model
  id_list <- strsplit(id, split = "[.]")[[1]]
  res <- rbind(list(Country = id_list[[1]], 
                    flavor = NA,
                    energy = "Without energy", 
                    model = "exp", 
                    nest = NA,
                    cst = "Reject CST"))
  res <- cbind(res, naturalCoef(modExp))
  # Now caluculate mean squared error (mse) for each model.
  sqResids <- resid(modExp)^2
  mse <- sum(sqResids) / (length(sqResids))
  res <- cbind(res, list(mse=mse))
  return(res)
}) %>% do.call(plyr::rbind.fill, .)

@@

Both historical GDP and fitted GDP are shown in 
Figures~\ref{fig:Graphs CD-fitted}--\ref{fig:Graphs CESekl-fitted}.
%
<<Graphs CD-fitted, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=4, fig.cap="Cobb-Douglas models that reject the cost-share theorem. Solid line is historical GDP.">>=
fittedGraphNoCST(filter(fitted_and_resid_data, model=="CD" & cst=="Reject CST"))
@
%
<<Graphs CESkle-fitted, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=4, fig.cap="CES models with ($kl$)$e$ nesting. Solid line is historical GDP.">>=
fittedGraphNoCST(filter(fitted_and_resid_data, model=="CES" & cst=="Reject CST" & (nest=="kle" | nest=="kl")))
@
%
<<Graphs CESlek-fitted, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=4, fig.cap="CES models with ($le$)$k$ nesting. Solid line is historical GDP.">>=
fittedGraphNoCST(filter(fitted_and_resid_data, model=="CES" & cst=="Reject CST" & (nest=="lek" | nest=="kl")))
@
%
<<Graphs CESekl-fitted, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=4, fig.cap="CES models with ($ek$)$l$ nesting. Solid line is historical GDP.">>=
fittedGraphNoCST(filter(fitted_and_resid_data, model=="CES" & cst=="Reject CST" & (nest=="ekl" | nest=="kl")))
@

@@

Figures~\ref{fig:Graphs CD-resid}--\ref{fig:Graphs CESekl-resid}
show fitting residuals for all models.
%
<<Graphs CD-resid, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=4, fig.cap="Fitting residuals for Cobb-Douglas models.">>=
residualPlot(filter(fitted_and_resid_data, model=="CD" & cst=="Reject CST"))
@
%
<<Graphs CESkle-resid, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=4, fig.cap="Fitting residuals for CES models with $(kl$)$e$ nesting.">>=
residualPlot(filter(fitted_and_resid_data, model=="CES" & cst=="Reject CST" & (nest=="kle" | nest=="kl")))
@
%
<<Graphs CESlek-resid, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=4, fig.cap="Fitting residuals for CES models with ($le$)$k$ nesting.">>=
residualPlot(filter(fitted_and_resid_data, model=="CES" & cst=="Reject CST" & (nest == "lek" | nest=="kl")))
@
%
<<Graphs CESekl-resid, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=4, fig.cap="Fitting residuals for CES models with $(ek$)$l$ nesting.">>=
residualPlot(filter(fitted_and_resid_data, model=="CES" & cst=="Reject CST" & (nest=="ekl" | nest=="kl")))
@

@@

Table~\ref{tab:Effect_of_CST_CD} shows the effect of the CST 
on fitted parameters for the Cobb-Douglas model and the exponential-only model.
The Cobb-Douglas models show lower $mse$ than the exponential-only models.
The Cobb-Douglas models also reduce solow residual ($\lambda$) relative
to the exponential-only case.
%
<<Effect_of_CST_CD, results="asis">>=
myXTable(filter(AllCoeffs, (model=="CD" & flavor=="Unadjusted" & energy=="Without energy") | model=="exp") %>% 
           select(Country, model, cst, scale, lambda, alpha_1, alpha_2, mse),
         caption="Model parameters for Cobb-Douglas models with unadjusted factors of production, without energy.",
         label=paste0("tab:", opts_current$get("label")), 
         digits=6)
@
%
Table~\ref{tab:Effect_of_CST_CES} shows the effect of the CST 
on fitted parameters for the CES model.
The CES models show lower $mse$ than the exponential-only models.
The CES models also reduce solow residual ($\lambda$) relative
to the exponential-only case.
%
<<Effect_of_CST_CES, results="asis">>=
myXTable(filter(AllCoeffs, (model=="CES" & flavor=="Unadjusted" & energy=="Without energy") | model=="exp") %>% 
           select(Country, model, cst, gamma, lambda, alpha_1, alpha_2, sigma_1, mse),
         caption="Model parameters for CES models with unadjusted factors of production, without energy.",
         label=paste0("tab:", opts_current$get("label")), 
         digits=6)
@
%
Both Cobb-Douglas and CES models do a better job of fitting 
historical data than an exponential-only model
as evidenced by decreased $mse$ values.
Both Cobb-Douglas and CES models exhibit reduced Solow residual~($\lambda$
compared to the exponential-only model.
For both Cobb-Douglas and CES, rejecting the Cost Share Theorem yields
decreased mean squared error~($mse$) and
smaller Solow residual ($\lambda$, on an absolute-value basis),
as expected.

@@

To calculate $\Delta \lambda$ and $\Delta mse$, we need to fit all combinations of 
country, model, flavor, energy, nest, and CST.
Table~\ref{tab:CD_coeffs} shows coefficients for all fitted Cobb-Douglas models
and the reference model (exponential-only).
<<CD_coeffs, results="asis">>=
myXTable(filter(AllCoeffs, model=="CD" | model=="exp") %>% 
           select(Country, model, flavor, energy, cst, scale, lambda, alpha_1, alpha_2, alpha_3, mse),
         caption="Model parameters for all Cobb-Douglas models.",
         label=paste0("tab:", opts_current$get("label")), 
         digits=6)
@
Tables~\ref{tab:CES_coeffs} and~\ref{tab:CES_coeffs_sigmas} show 
coefficients for all fitted CES models and the reference model (exponential-only).
<<CES_coeffs, results="asis">>=
myXTable(filter(AllCoeffs, model=="CES" | model=="exp") %>% 
           select(Country, model, flavor, energy, nest, cst, gamma, lambda, alpha_1, alpha_2, alpha_3),
         caption="Model parameters for all CES models.",
         label=paste0("tab:", opts_current$get("label")), 
         digits=6)
@
<<CES_coeffs_sigmas, results="asis">>=
myXTable(filter(AllCoeffs, model=="CES" | model=="exp") %>% 
           select(Country, model, flavor, energy, nest, cst, sigma_1, sigma, mse),
         caption="Model parameters for all CES models.",
         label=paste0("tab:", opts_current$get("label")), 
         digits=6)
@
Table~\ref{tab:DSRmse_CD} shows $\Delta \lambda$ and $\Delta mse$ results for the Cobb-Douglas models.
<<DSRmse_CD, results="asis">>=
myXTable(filter(AllCoeffs, model=="CD") %>% select(Country, model, flavor, energy, cst, Dlambda, Dmse),
         caption="$\\Delta \\lambda$ and $\\Delta mse$ for Cobb-Douglas models.",
         label=paste0("tab:", opts_current$get("label")), 
         digits=6)
@
Tables~\ref{tab:DSRmse_CESkle}--\ref{tab:DSRmse_CESekl} show $\Delta \lambda$ and $\Delta mse$ 
results for CES models with various nestings.
<<DSRmse_CESkle, results="asis">>=
myXTable(filter(AllCoeffs, (model=="CES" & (nest=="kl" | nest=="kle")) | model=="exp") %>% 
           select(Country, model, flavor, energy, nest, cst, Dlambda, Dmse),
         caption="$\\Delta \\lambda$ and $\\Delta mse$ for CES models with ($kl$)$e$ nesting.",
         label=paste0("tab:", opts_current$get("label")), 
         digits=6)
@
<<DSRmse_CESlek, results="asis">>=
myXTable(filter(AllCoeffs, (model=="CES" & (nest=="kl" | nest=="lek")) | model=="exp") %>% 
           select(Country, model, flavor, energy, nest, cst, Dlambda, Dmse),
         caption="$\\Delta \\lambda$ and $\\Delta mse$ for CES models with ($le$)$k$ nesting.",
         label=paste0("tab:", opts_current$get("label")), 
         digits=6)
@
<<DSRmse_CESekl, results="asis">>=
myXTable(filter(AllCoeffs, (model=="CES" & (nest=="kl" | nest=="ekl")) | model=="exp") %>% 
           select(Country, model, flavor, energy, nest, cst, Dlambda, Dmse),
         caption="$\\Delta \\lambda$ and $\\Delta mse$ for CES models with ($ek$)$l$ nesting.",
         label=paste0("tab:", opts_current$get("label")), 
         digits=6)
@
Figures~\ref{fig:DSRmse_CD_graph} and~\ref{fig:percDSRmse_CD_graph} summarize 
$\Delta \lambda$ and $\Delta mse$ 
results for the Cobb-Douglas model.
<<DSRmse_CD_graph, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=4, fig.cap="Change in Solow Residuals~($\\Delta\\lambda$) and mean squared error~($\\Delta mse$) for the Cobb-Douglas model relative to the exponential-only model.">>=
ggplot() +
  geom_point(data=filter(AllCoeffs, model=="CD"), mapping=aes(x=Dmse, y=Dlambda, color=energy, size=cst)) +
  facet_grid(Country ~ flavor) +
  scale_size_manual(values=c(2,3.5)) +
  xlab(expression(Delta * mse)) + 
  scale_x_continuous(breaks=seq(-0.01, 0, 0.005), limits=c(-0.012, 0.003)) +  
  ylab(expression(Delta * lambda)) +
  scale_y_continuous(limits=c(-0.04, 0)) +
  xy_theme() + 
  theme(legend.title=element_blank())
@
<<percDSRmse_CD_graph, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=4, fig.cap="Percentage change in Solow Residuals~($\\Delta\\lambda \\, [\\%]$) and mean squared error~($\\Delta mse \\, [\\%]$) for the Cobb-Douglas model relative to exponential-only models.">>=
ggplot() +
  geom_point(data=filter(AllCoeffs, model=="CD"), mapping=aes(x=percDmse, y=percDlambda, color=energy, size=cst)) +
  facet_grid(Country ~ flavor) +
  scale_size_manual(values=c(2,3.5)) +
  xlab(expression(paste(Delta * mse, " [%]"))) + 
  scale_x_continuous(breaks=seq(-50, 50, 50), limits=c(-90, 50)) +  
  ylab(expression(paste(Delta * lambda, " [%]"))) +
  scale_y_continuous(breaks=seq(-80, 0, 20), limits=c(-100, 10)) +  
  xy_theme() + 
  theme(legend.title=element_blank())
@
Figures~\ref{fig:DSRmse_CES_graph} and~\ref{fig:percDSRmse_CES_graph} summarize 
$\Delta \lambda$ and $\Delta mse$ 
results for CES models.
<<DSRmse_CES_graph, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=4, fig.cap="Change in Solow Residuals~($\\Delta\\lambda$) and mean squared error~($\\Delta mse$) for the CES modles relative to the exponential-only model.">>=
ggplot() +
  geom_point(data=filter(AllCoeffs, model=="CES"), mapping=aes(x=Dmse, y=Dlambda, color=energy, size=cst, shape=nest)) +
  scale_size_manual(values=c(2,3.5)) +
  scale_shape_manual(values=c(16,17,18,15)) +
  facet_grid(Country ~ flavor) +
  xlab(expression(Delta * mse)) + 
  scale_x_continuous(breaks=seq(-0.005, 0, 0.005)) +  
  ylab(expression(Delta * lambda)) +
  xy_theme() +
  theme(legend.title=element_blank())
@
<<percDSRmse_CES_graph, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=4, fig.cap="Percentage change in Solow Residuals~($\\Delta\\lambda \\, [\\%]$) and mean squared error~($\\Delta mse \\, [\\%]$) for the CES models relative to the exponential-only model.">>=
ggplot() +
  geom_point(data=filter(AllCoeffs, model=="CES"), mapping=aes(x=percDmse, y=percDlambda, color=energy, size=cst, shape=nest)) +
  scale_size_manual(values=c(2,3.5)) +
  scale_shape_manual(values=c(16,17,18,15)) +
  facet_grid(Country ~ flavor) +
  xlab(expression(paste(Delta * mse, " [%]"))) + 
  scale_x_continuous(breaks=seq(-80, 0, 20)) +  
  ylab(expression(paste(Delta * lambda, " [%]"))) +
  scale_y_continuous(breaks=seq(-80, 0, 20), limits=c(-100, 10)) +  
  xy_theme() +
  theme(legend.title=element_blank())
@

@@

The set of fitting residual values~($r_i$)
for each combination of 
country~(PT or UK), 
model~(Cobb-Douglas or CES),
flavor~(Unadjusted or Quality-adjusted), 
energy~(with or without), and
nest~[($kl$), ($kl$)$e$, ($le$)$k$, or ($ek$)$l$, for the CES models]
can be assessed for statistical significance
compared to zero by a t-test.
The null hypothesis for the t-test is that the mean of each group is
equal to zero, and 
the alternative hypothesis is that the mean is not equal to zero.
Smaller p-values indicate increasing confidence that the 
mean is different from zero. 
We expect that p-values will be large (close to unity), 
because the mean of the residuals is expected to be small (close to zero).
Indeed, that is the case.
Table~\ref{tab:r_p_values} summarizes the mean 
and associated p-values values for the fitting residuals~($r_i$).
%
<<r_p_values, results="asis">>=
# Calculate p values for groups of fitting residuals.
r.t.values <- fitted_and_resid_data %>% 
  group_by(Country, model, flavor, energy, nest) %>% 
  summarize(Residuals.mean = mean(Residuals), Residuals.p.value = stats::t.test(Residuals)[["p.value"]])
r.t.values <- arrange(ungroup(r.t.values), Country, model, flavor, energy, nest)
# Display the values in a table
myXTable(r.t.values,
         caption="Means and p-values for fitting residuals~($r_i$).",
         label=paste0("tab:", opts_current$get("label")), 
         digits=6)
@

@@

The set of annual $\Delta r$ values for each combination of 
country~(PT or UK), 
model~(Cobb-Douglas or CES),
flavor~(Unadjusted or Quality-adjusted), 
energy~(with or without), and
nest~[($kl$), ($kl$)$e$, ($le$)$k$, or ($ek$)$l$, for the CES models]
can be assessed for statistical significance
compared to zero by a t-test.
The null hypothesis for the t-test is that the mean of each group is
equal to zero, and 
the alternative hypothesis is that the mean is less than zero.
Smaller p-values indicate increasing confidence that the 
mean is actually less than zero.
Table~\ref{tab:Dr_p_values} summarizes the mean 
and associated p-values values for $\Delta r$.
%
<<Dr_p_values, results="asis">>=
# Calculate p values for groups of Dr values.
Dr.t.values <- fitted_and_resid_data %>% 
  group_by(Country, model, flavor, energy, nest) %>% 
  summarize(Dr.mean = mean(Dr), Dr.p.value = stats::t.test(Dr, alternative = "less")[["p.value"]])
Dr.t.values <- arrange(ungroup(Dr.t.values), Country, model, flavor, energy, nest)
# Display the values in a table
myXTable(Dr.t.values,
         caption="Means and p-values for $\\Delta r$.",
         label=paste0("tab:", opts_current$get("label")), 
         digits=6)
@

>>>>

# Tell whether to re-run the analysis to create the models.
# This file will use previously-saved models to generate
# the paper if set to FALSE, greatly reducing time.
# If something changes in the data or in the code that
# performs the analysis, set 
# rerunModels = TRUE
# once to re-generate the information needed to 
# create the paper.
# ********************************************************
rerunModels=FALSE
# ********************************************************
opts_chunk$set(
  # ********************************************************
  eval=TRUE,
  # ********************************************************
  # tikz allows LaTeX code in graphical output. 
  # E.g., "$y$" for a variable name in the legend for a graph.  
  # dev='tikz',
  tidy=FALSE,
  comment=NA,
  # Tells whether to cache output from chunks are saved. Cacheing saves time. 
  # However, references to LaTeX tables DO NOT WORK if cacheing is turned on.
  cache=FALSE,
  warning=FALSE,  #Tells whether to show warnings in the output.
  # Turns off messages for all chunks. 
  # Set TRUE on an individual chunk to see it.
  message=FALSE, 
  # Tells whether to echo code for all chunks. 
  # Set TRUE on an individual chunk to see its code.
  echo=FALSE,
  # Sets all table captions to the top.
  xtable.caption.placement="top"
)

@@

%------------------------------
\subsubsection{Exponential Production Function (reference model)} 
\label{sec:exp}
%------------------------------

We define an exponential-only reference model for economic growth
in Equation~\ref{eq:exponential-only-model}.
%
\begin{equation} \label{eq:exponential-only-model}
  y = \theta A \; ; \; A \equiv \mathrm{e}^{\lambda(t-t_0)} \; .
\end{equation}
%
The reference model is a degenerate case of the Cobb-Douglas production function
wherein the constant returns to scale constraint is not respected,
because all factor shares are zero~($\alpha_1 = \alpha_2 = \alpha_3 = 0$).

We expect that the reference model will have a larger fitted Solow residual term than
the Cobb-Douglas and CES models, because no factors of production are
included in the reference model to drive economic growth.
Indeed, in the reference model, 
all economics growth is attributed to the Solow residual.

@@

  models <- list()
  for(country in countries){
    for (flavor in flavors){
      fnoE <- keep_summands(formulas[[country]][[flavor]], -3) # Drop the energy term (3rd term on RHS)
      fwithE <- formulas[[country]][[flavor]]
      # CD models
      models[[country]][[flavor]][["noE"]][["CD"]]   <- cdModel(fnoE, data = historicalData[[country]])
      models[[country]][[flavor]][["withE"]][["CD"]] <- cdModel(fwithE, data = historicalData[[country]])
      # CES models
      models[[country]][[flavor]][["noE"]][["CES"]][["kl"]] <- cesModel(fnoE, 
                                                                        data = historicalData[[country]], 
                                                                        nest=c(1,2))
      nestsWithE <- list(kle=c(1,2,3), lek=c(2,3,1), ekl=c(3,1,2))
      models[[country]][[flavor]][["withE"]][["CES"]] <- lapply(nestsWithE, function(nest){
          cesModel(formula=fwithE, data=historicalData[[country]], nest=nest)
        })
    }      
  }
  # Save models to disk.
  saveRDS(models, file=fileNameCDCES)
}
# Put all exponential model coefficients into a data frame
AllCoeffsExp <- leaf_apply(modelsExp, class = c("lm"), f = function(modExp, id){
  # Parse the id to find details of this model
  id_list <- strsplit(id, split = "[.]")[[1]]
  res <- rbind(list(Country = id_list[[1]], 
                    flavor = NA,
                    energy = "Without energy", 
                    model = "exp", 
                    nest = NA,
                    cst = "Reject CST"))
  res <- cbind(res, naturalCoef(modExp))
  # Now caluculate mean squared error (mse) for each model.
  sqResids <- resid(modExp)^2
  mse <- sum(sqResids) / (length(sqResids))
  res <- cbind(res, list(mse=mse))
  return(res)
}) %>% do.call(plyr::rbind.fill, .)

@@

Both historical GDP and fitted GDP are shown in 
Figures~\ref{fig:Graphs CESkle-fitted}--\ref{fig:Graphs CESekl-fitted}.
%
<<Graphs CESkle-fitted, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=4, fig.cap="CES models with ($kl$)$e$ nesting. Solid line is historical GDP.">>=
fittedGraphNoCST(filter(fitted_and_resid_data, model=="CES" & cst=="Reject CST" & (nest=="kle" | nest=="kl")))
@
%
<<Graphs CESlek-fitted, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=4, fig.cap="CES models with ($le$)$k$ nesting. Solid line is historical GDP.">>=
fittedGraphNoCST(filter(fitted_and_resid_data, model=="CES" & cst=="Reject CST" & (nest=="lek" | nest=="kl")))
@
%
<<Graphs CESekl-fitted, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=4, fig.cap="CES models with ($ek$)$l$ nesting. Solid line is historical GDP.">>=
fittedGraphNoCST(filter(fitted_and_resid_data, model=="CES" & cst=="Reject CST" & (nest=="ekl" | nest=="kl")))
@

@@

Figures~\ref{fig:Graphs CESkle-resid}--\ref{fig:Graphs CESekl-resid}
show fitting residuals for all CES models.
%
<<Graphs CESkle-resid, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=4, fig.cap="Fitting residuals for CES models with $(kl$)$e$ nesting.">>=
residualPlot(filter(fitted_and_resid_data, model=="CES" & cst=="Reject CST" & (nest=="kle" | nest=="kl")))
@
%
<<Graphs CESlek-resid, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=4, fig.cap="Fitting residuals for CES models with ($le$)$k$ nesting.">>=
residualPlot(filter(fitted_and_resid_data, model=="CES" & cst=="Reject CST" & (nest == "lek" | nest=="kl")))
@
%
<<Graphs CESekl-resid, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=4, fig.cap="Fitting residuals for CES models with $(ek$)$l$ nesting.">>=
residualPlot(filter(fitted_and_resid_data, model=="CES" & cst=="Reject CST" & (nest=="ekl" | nest=="kl")))
@

The primary results from this paper involve the effects of
the following modeling decisions:
rejecting (or adhering to) the cost share theorem,
quality-adjusting (or not) the factors of production,
including energy (or not) in the production function, and
nesting (in the CES model).

We evaluate the effects of these modeling decisions 
first on goodness of fit and Solow residual,
next on factor shares, and 
finally on factor substitutability.

%++++++++++++++++++++++++++++++
\subsection{Goodness of fit and Solow residual} 
\label{sec:fit_and_sr}
%++++++++++++++++++++++++++++++

%------------------------------
\subsubsection{Rejecting (or adhering to) the cost share theorem} 
\label{sec:gfsr_cst}
%------------------------------

%------------------------------
\subsubsection{Quality-adjusting (or not) factors of production} 
\label{sec:gfsr_fop}
%------------------------------

%------------------------------
\subsubsection{Including energy (or not)} 
\label{sec:gfsr_energy}
%------------------------------

%------------------------------
\subsubsection{CES nesting} 
\label{sec:gfsr_nesting}
%------------------------------

%++++++++++++++++++++++++++++++
\subsection{Factor shares} 
\label{sec:shares}
%++++++++++++++++++++++++++++++

%------------------------------
\subsubsection{Rejecting (or adhering to) the cost share theorem} 
\label{sec:fs_cst}
%------------------------------

%------------------------------
\subsubsection{Quality-adjusting (or not) factors of production} 
\label{sec:fs_fop}
%------------------------------

%------------------------------
\subsubsection{Including energy (or not)} 
\label{sec:fs_energy}
%------------------------------

%------------------------------
\subsubsection{CES nesting} 
\label{sec:fs_nesting}
%------------------------------

%++++++++++++++++++++++++++++++
\subsection{Factor substitutability} 
\label{sec:substitutability}
%++++++++++++++++++++++++++++++

%------------------------------
\subsubsection{Rejecting (or adhering to) the cost share theorem} 
\label{sec:sigma_cst}
%------------------------------

%------------------------------
\subsubsection{Quality-adjusting (or not) factors of production} 
\label{sec:sigma_fop}
%------------------------------

%------------------------------
\subsubsection{Including energy (or not)} 
\label{sec:sigma_energy}
%------------------------------

%------------------------------
\subsubsection{CES nesting} 
\label{sec:sigma_nesting}
%------------------------------

@@

Table~\ref{tab:Effect_of_CST_CES} shows the effect of the CST 
on fitted parameters for the CES model.
The CES models show lower $mse$ than the exponential-only models.
The CES models also reduce solow residual ($\lambda$) relative
to the exponential-only case.
%
<<Effect_of_CST_CES, results="asis">>=
myXTable(filter(AllCoeffs, (model=="CES" & flavor=="Unadjusted" & energy=="Without energy") | model=="exp") %>% 
           select(Country, model, cst, gamma, lambda, alpha_1, alpha_2, sigma_1, mse),
         caption="Model parameters for CES models with unadjusted factors of production, without energy.",
         label=paste0("tab:", opts_current$get("label")), 
         digits=6)
@
%
Both Cobb-Douglas and CES models do a better job of fitting 
historical data than an exponential-only model
as evidenced by decreased $mse$ values.
Both Cobb-Douglas and CES models exhibit reduced Solow residual~($\lambda$
compared to the exponential-only model.
For both Cobb-Douglas and CES, rejecting the Cost Share Theorem yields
decreased mean squared error~($mse$) and
smaller Solow residual ($\lambda$, on an absolute-value basis),
as expected.

@@

To calculate $\Delta \lambda$ and $\Delta mse$, we need to fit all combinations of 
country, model, flavor, energy, nest, and CST.
Tables~\ref{tab:CES_coeffs} and~\ref{tab:CES_coeffs_sigmas} show 
coefficients for all fitted CES models and the reference model (exponential-only).
%
<<CES_coeffs, results="asis">>=
myXTable(filter(AllCoeffs, model=="CES" | model=="exp") %>% 
           select(Country, model, flavor, energy, nest, cst, gamma, lambda, alpha_1, alpha_2, alpha_3),
         caption="Model parameters for all CES models.",
         label=paste0("tab:", opts_current$get("label")), 
         digits=6)
@
%
<<CES_coeffs_sigmas, results="asis">>=
myXTable(filter(AllCoeffs, model=="CES" | model=="exp") %>% 
           select(Country, model, flavor, energy, nest, cst, sigma_1, sigma, mse),
         caption="Model parameters for all CES models.",
         label=paste0("tab:", opts_current$get("label")), 
         digits=6)
@
%
Tables~\ref{tab:DSRmse_CESkle}--\ref{tab:DSRmse_CESekl} show $\Delta \lambda$ and $\Delta mse$ 
results for CES models with various nestings.
%
<<DSRmse_CESkle, results="asis">>=
myXTable(filter(AllCoeffs, (model=="CES" & (nest=="kl" | nest=="kle")) | model=="exp") %>% 
           select(Country, model, flavor, energy, nest, cst, Dlambda, Dmse),
         caption="$\\Delta \\lambda$ and $\\Delta mse$ for CES models with ($kl$)$e$ nesting.",
         label=paste0("tab:", opts_current$get("label")), 
         digits=6)
@
%
<<DSRmse_CESlek, results="asis">>=
myXTable(filter(AllCoeffs, (model=="CES" & (nest=="kl" | nest=="lek")) | model=="exp") %>% 
           select(Country, model, flavor, energy, nest, cst, Dlambda, Dmse),
         caption="$\\Delta \\lambda$ and $\\Delta mse$ for CES models with ($le$)$k$ nesting.",
         label=paste0("tab:", opts_current$get("label")), 
         digits=6)
@
%
<<DSRmse_CESekl, results="asis">>=
myXTable(filter(AllCoeffs, (model=="CES" & (nest=="kl" | nest=="ekl")) | model=="exp") %>% 
           select(Country, model, flavor, energy, nest, cst, Dlambda, Dmse),
         caption="$\\Delta \\lambda$ and $\\Delta mse$ for CES models with ($ek$)$l$ nesting.",
         label=paste0("tab:", opts_current$get("label")), 
         digits=6)
@
Figures~\ref{fig:DSRmse_CES_graph} and~\ref{fig:percDSRmse_CES_graph} summarize 
$\Delta \lambda$ and $\Delta mse$ 
results for CES models.
%
<<DSRmse_CES_graph, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=4, fig.cap="Change in Solow Residuals~($\\Delta\\lambda$) and mean squared error~($\\Delta mse$) for the CES modles relative to the exponential-only model.">>=
ggplot() +
  geom_point(data=filter(AllCoeffs, model=="CES"), mapping=aes(x=Dmse, y=Dlambda, color=energy, size=cst, shape=nest)) +
  scale_size_manual(values=c(2,3.5)) +
  scale_shape_manual(values=c(16,17,18,15)) +
  facet_grid(Country ~ flavor) +
  xlab(expression(Delta * mse)) + 
  scale_x_continuous(breaks=seq(-0.005, 0, 0.005)) +  
  ylab(expression(Delta * lambda)) +
  xy_theme() +
  theme(legend.title=element_blank())
@
%
<<percDSRmse_CES_graph, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=4, fig.cap="Percentage change in Solow Residuals~($\\Delta\\lambda \\, [\\%]$) and mean squared error~($\\Delta mse \\, [\\%]$) for the CES models relative to the exponential-only model.">>=
ggplot() +
  geom_point(data=filter(AllCoeffs, model=="CES"), mapping=aes(x=percDmse, y=percDlambda, color=energy, size=cst, shape=nest)) +
  scale_size_manual(values=c(2,3.5)) +
  scale_shape_manual(values=c(16,17,18,15)) +
  facet_grid(Country ~ flavor) +
  xlab(expression(paste(Delta * mse, " [%]"))) + 
  scale_x_continuous(breaks=seq(-80, 0, 20)) +  
  ylab(expression(paste(Delta * lambda, " [%]"))) +
  scale_y_continuous(breaks=seq(-80, 0, 20), limits=c(-100, 10)) +  
  xy_theme() +
  theme(legend.title=element_blank())
@

@@

The set of fitting residual values~($r_i$)
for each combination of 
country~(PT or UK), 
model~(Cobb-Douglas or CES),
flavor~(Unadjusted or Quality-adjusted), 
energy~(with or without), and
nest~[($kl$), ($kl$)$e$, ($le$)$k$, or ($ek$)$l$, for the CES models]
can be assessed for statistical significance
compared to zero by a t-test.
The null hypothesis for the t-test is that the mean of each group is
equal to zero, and 
the alternative hypothesis is that the mean is not equal to zero.
Smaller p-values indicate increasing confidence that the 
mean is different from zero. 
We expect that p-values will be large (close to unity), 
because the mean of the residuals is expected to be small (close to zero).
Indeed, that is the case.
Table~\ref{tab:r_p_values} summarizes the mean 
and associated p-values values for the fitting residuals~($r_i$).
%
<<r_p_values, results="asis">>=
# Calculate p values for groups of fitting residuals.
r.t.values <- filter(fitted_and_resid_data, model != "CD") %>% 
  group_by(Country, model, flavor, energy, nest) %>% 
  summarize(Residuals.mean = mean(Residuals), Residuals.p.value = stats::t.test(Residuals)[["p.value"]])
r.t.values <- arrange(ungroup(r.t.values), Country, model, flavor, energy, nest)
# Display the values in a table
myXTable(r.t.values,
         caption="Means and p-values for fitting residuals~($r_i$).",
         label=paste0("tab:", opts_current$get("label")), 
         digits=6)
@

@@

The set of annual $\Delta r$ values for each combination of 
country~(PT or UK), 
model~(Cobb-Douglas or CES),
flavor~(Unadjusted or Quality-adjusted), 
energy~(with or without), and
nest~[($kl$), ($kl$)$e$, ($le$)$k$, or ($ek$)$l$, for the CES models]
can be assessed for statistical significance
compared to zero by a t-test.
The null hypothesis for the t-test is that the mean of each group is
equal to zero, and 
the alternative hypothesis is that the mean is less than zero.
Smaller p-values indicate increasing confidence that the 
mean is actually less than zero.
Table~\ref{tab:Dr_p_values} summarizes the mean 
and associated p-values values for $\Delta r$.
%
<<Dr_p_values, results="asis">>=
# Calculate p values for groups of Dr values.
Dr.t.values <- filter(fitted_and_resid_data, model != "CD") %>% 
  group_by(Country, model, flavor, energy, nest) %>% 
  summarize(Dr.mean = mean(Dr), Dr.p.value = stats::t.test(Dr, alternative = "less")[["p.value"]])
Dr.t.values <- arrange(ungroup(Dr.t.values), Country, model, flavor, energy, nest)
# Display the values in a table
myXTable(Dr.t.values,
         caption="Means and p-values for $\\Delta r$.",
         label=paste0("tab:", opts_current$get("label")), 
         digits=6)
@
